\documentstyle[ltugbot,crossword]{article}
\setcounter{page}{324}
\SelfDocumenting
\PrelimDraftfalse
\begin{document}
\vspace*{-20pt}%  This is horrible, but it's the only way I could fit the
               %  whole crossword and the inroductory text onto the page.
	       %  If you can find a better way, please use it!
\begin{crossword}{15}{N}
%\centerline{\bf No.~1\hfill by Logo\TeX nites}
\centerline{\bf No.~1\hfill by Logo\TeX nes}
{\scriptsize % Use of \scriptsize is also pretty horrible!
Please note that the somewhat insulting clues for certain
personages are purely a compiler's artifice!\\
(One American spelling, and certain proper nouns are not in {\it Chambers}.)\\}
% Across clues
\clue{5}{A}{10}{1}{SUFFIX}{Sextet bring you in to play very loudly
underscore introduction}{6}
\clue{8}{A}{1}{2}{ACADEMIA}{A metallic element media disturbed in halls of
learning}{8} 
\clue{9}{A}{9}{3}{BRACKET}{}{see 20}
\clue{10}{A}{1}{4}{ALPHA}{Letter from Greek mountain area}{5}
\clue{11}{A}{7}{5}{MAGNESITE}{Girl with sex appeal in Middle Eastern source
of silvery metal}{9}
\clue{13}{A}{1}{6}{SCREENED}{Senate decrees North will be made to give a
public exhibition}{8}
\clue{14}{A}{10}{7}{EGRETS}{Sorrowful wishes don't begin to describe white
herons}{6} 
\clue{17}{A}{5}{8}{END}{Finally pairs up with 28!}{3}
\clue{19}{A}{9}{8}{GNU}{Sounds like a new companion for Nat}{3}
\clue{20 \noexpand\& 9}{A}{1}{9}{SQUARE}{Eastern quarter-backs provide
optional introduction}{6,7}
\clue{23}{A}{8}{10}{ASTEROID}{A body-builder produces heavenly body!}{8}
\clue{26}{A}{1}{11}{BACKSLASH}{Without this character's help, there's no
escape!}{9}
\clue{28}{A}{11}{12}{BEGIN}{Make a start on fencing in the environment?}{5}
\clue{29}{A}{1}{13}{MINUEND}{Note taken from short time to finish? No,
but that from which it's taken}{7}
\clue{30}{A}{8}{14}{OVERFULL}{Scour and beat after concerning boxes which
are too high or wide}{8}
\clue{31}{A}{1}{15}{INCHES}{Scots island points after Imperial measures}{6}
% Down clues
\clue{1}{D}{1}{1}{FARADS}{Measures of capacity formerly unused by 27}{6}
\clue{2}{D}{3}{1}{LAMPORT}{Too bad French brought up this man's lack of
multilingual support}{7}
\clue{3}{D}{5}{1}{DEKAMETER}{1,864,679,811 of the smallest units 27
abbreviated a measurer}{9}
\clue{4}{D}{7}{1}{BIREME}{Galley filled with two lines of Roman types?
Rim involved six-footer in its construction}{6}
\clue{5}{D}{10}{1}{SHRUNKEN}{Shun kern construction: it's contracted}{8}
\clue{6}{D}{12}{1}{FUCHS}{This man spreads \noexpand\TeX's secrets like his
atomic namesake!}{5} 
\clue{7}{D}{14}{1}{IDENTITY}{Tiny diet indisposed personality}{8}
\clue{12}{D}{8}{5}{ADA}{Very refined language used by world's first
programmer?}{3} 
\clue{15}{D}{11}{7}{GUTENBERG}{Moveable type, inventor gent moved around
eastern burg}{9}
\clue{16}{D}{2}{8}{EQUATION}{One mixed up about insignificant Shakespearian
person with another one, yields formula to express chemical action}{8}
\clue{18}{D}{6}{8}{NEWLINES}{Fresh verses: two or more finish the paragraph}{8}
\clue{21}{D}{8}{9}{GAS}{Talk a lot of hot air!}{3}
\clue{22}{D}{13}{9}{SORGHUM}{Chewing sort of sugar cane could lead to tender
oral tissue, we hear}{7}
\clue{24}{D}{9}{10}{SHEAVE}{Woman greeting a speck in paper}{6}
\clue{25 \noexpand\& 27}{D}{15}{10}{DONALD}{Hunt odd lank confused
professor}{6,5} 
\clue{27}{D}{4}{11}{KNUTH}{}{}
\end{crossword}
\end{document}
