\input eplain

\tolerance=2000

\def\SP{\ }
\def\throwaway#1{}
\def\checkspace{\ifx \next\
\let\result=\throwaway
 \else \let\result=\ignorespaces \fi \result }
\font\bigfont=cmr10 scaled \magstep5
\def\bfl#1{\setbox0=\hbox {{\bigfont #1}\thinspace}
      \noindent \hangindent \wd0 \hangafter -2
      \llap {\vbox to \ht0{\kern\baselineskip \box0 \kern-\baselineskip}}%
      \futurelet \next \checkspace}

\parskip=10pt
\hbadness = 1500
\parindent=0pt

\font\cmbig=cmr17
\font\cmrbv=cmb10 scaled \magstep5
\font\iti=cmbxti10 scaled \magstep2
\font\erm=cmr8

\doublecolumns

%**************************************************************************
% Generate a dummy first page for the twocol macro
% The first page is to have the left column blank.

\phantom{dummy first line} \vskip5truein
 {\font\iti=cmti10 scaled \magstep1
 \iti
 \rightline{An abstract}
 \rightline{goes}
 \rightline{in here.}
 }

%karl \vfill
%karl \eject

%**************************************************************************

\phantom{dummy first line} \vskip0.75truein

\rightline{\cmrbv Title}
\bigskip\bigskip
\rightline{\iti Author}

\bigskip\bigskip\bigskip
%\rightline{\hrulefill}% width6.5truein}

\nointerlineskip
\moveleft 3.31truein
\vbox{\hrule width6.5truein}
\nointerlineskip

\vskip0.75truein

%**************************************************************************

\bfl
Xxxxxxx xxxxxx xxxxxx xxxxxxx xxxx xxx xxx\numberedfootnote{\erm Xxxxx xxxxx
xxx xxxx Xxxxx xxxxx, Xxxxx xxxx Xxxxxx xxxxxx, xxxxxx xxxxx xx Xxxxxx Xxxxxx,
Xxx., X.~X. Xxx~xxx, Xxxx Xxxxx, XX~xxxxx.} Xxxx Xxxxxxxxxx xxxxxxxx xxxx, Xxx
xxxxx xxxxxx xx xxxxxxxx xx xxxxx xxxxxxx xxxxxxx xxxx xx xxxx xx xxxxxxx xx
xxx xxxxx xx x xxx xxxxxxx xx xxxxx. Xx x/xxx xx x xxx xxxx xxxxxxx xxxxx
xxxxxxx, Xxxxxx xxxxxxxxx xxxxx xxxxx x/xxx,xxx,xxx xx x xxxxxx. Xxxx xxxxx
 xxxx xx xxxx xxxx Xxxxxx'x xxxxxxxx xx xxx xxxxxxx xxxxxxxx xxxxxx xx xxxxx
 xx xxxx xxxxxxx xxx xxxxxx xxxx xx xxxxxxx xx xxxxxx x.xx xx x xxx, xxx x.xx
xxx xxxxxx xxxxx xx~xxxxxxx. Xxxx xx x xxxxx xxxxxx xxx xxxxxxxxx xxxxxxxx xxx
xxxxxxxx xxxx xxxxxxxx xxxxxxxxx xxx xxxxxx xxxxxx xxxx, xxx xxx xxxxxxxx
xxxxxxxxx xx xxx xxxx xx xxxxxxxxxx xx x xxxx xxxxx xxxxxxxxxx.

Xxxxx xxx xxxxxx xxx xxxxxx xx xxx xxxxx xx xxx xxx xxxxxxxxx (xxx XX-xxx
xxxxx xxxxxx xxxxxxxx xx xxx Xxx Xxxxxxxxxx Xxxxxxxxxx), xx xxx xxxxxxx xxxx
xx xxxxx xx xxxxxxxxxxx xx xxxxxxx x xxxxx xxxxxxx xxx xxxx xx xxxxx xxxx
xxxxxx xx xxxx x xxxxxxxxxxx xxxxxx xxxx xxx xxxxxxx.

Xxxxx x xxxxx xx xxx xxxx xx xxxxxxxxxx xxx xxxxxxxx xx xxx xxxxxxxxxx xxx
xxxxxx xxxxxxxx xxx xxxx xxxxxx xxx xxxxx xxxx. Xxxx xxx xxxx xxxx xxxx xx
xxxxx xxx xxxxxx'x xxxx xxxx xxxxxxxx, xxxx {xxx xxxxxxx} xx xxx xxxxxxx
xxxxxxxxx, xxxxxxxx, xx x, xxx xxxxxxxx. Xxxxx xxxx xxxxxxxxx xxx xxxx (xxxx
xxx Xxxx) xxx xxxxx xx Xxxxx~x.

Xx xxxxxxxxxxx xx Xxxxx~x xxxxxxx xxxx xxx xxxxxxxx xx xxxxxxx xx xxxx xxxxxxx
xxxxxxxx xxx xxx xx xxxxxxxxxx xx xxxxx xx xxxxxxxx. Xx xxxx, xx xxxxxxx xx
xxxxxx xxx xxxxx xx xx xxxxx xxx xxx xxxxxxxxx xxx xxxxxxxx xxxxx xx Xxxxxx'x
xxxxx xx Xxxxxx' xxxxxx xxxxxxx. Xxxx xxx xxxxxxxx xxxxx xxx xx xx xxxx, xxx
xxxxxxxxx. Xxxxx~x xxxxxxxx xxx xxxxxxxxx xx xxxxxxx xx xxx xxxxxx xxxxxx xx
xxxx xxxxxx xxxxxxxx xx xxxxxxxxx xx xxxxxxxxx.


\singlecolumn
\centerline{Table 1---Xxxxxxxxx Xxxx Xxxxxxxx xx Xxxxxxxx}
\smallbreak
\line \bgroup \hss
  \vbox to 11\baselineskip{}
\hss \egroup
\doublecolumns


Xxxx xxxxxx x xxxxxx xxxxxxx Xxxxxx'x xxxxxx xxxxxx xx xxx xxxxxx xxxxx xxxxxx
xxx xxxx xxxxx xx Xxxxx~x, xxxx xxxx xxxxxx xx xxxxx xxxxxxx. Xx, xxx xxxxxxx,
xx xxxxxxxxxx xxx xxxxxxxx xxxx xxxxxx xxxxxx, xx xxxxx xxxxx xx xxxxxxxxxxx,
xxxxxxxxxx, xxx xxxxxxxxx, xxxxxxx xx xxx xxxxxxxx xx xxxx xxxxxx xxxx xxxxx
xxxxxxx, xx xxxxx xxxxx xx xxx xxxxxxxx, xxxxx xxxxx, xxx xxxxxxxxx. Xxxx xxx
xxxx xx xxxxxxxxx.

\medskip\bigskip
\centerline{Table 2---Xxxxxxx xx Xxxxxx'x Xxxxxx Xxxxxxx}
\nobreak
\centerline{(Xxxxxxx Xxxxxxxx)}
\smallbreak
\line \bgroup \hss
  \vbox to 9\baselineskip{}
\hss \egroup

Xxxxxxxx xxx xxxx xxxx xxxxx xxxxx xx Xxxxx~x xxx xxx xxxx, xxxx xxx xxx xxxx
xx xxx xxxxxxxx xxxxxxxx xxxxxxxxx xxxxx. Xx xxx x xxxxx xxxxxxx xx xxx xxxx
xx xxxxxxx, xx xxx xxxxxxx xx xx x xxxxxx xxxxxx xx xxxxx xxxxxxx xx xxxxxx xx
xxxx xxxxx xxxxxx x ``xxxxx xxxx'' xxxxxxxx. Xxxxx x xxxxx xx xxxx xxxxx xx
xxx xxxx xx xxxxxxx xxx xxxxxxxx. Xxxx xxx xxxx xxxxxxxx xx xxxxx xxx xxxxxx'x
xxxxxxxx xxxxxx xx xxxxx xxxxx, xxxxxxx xxx, xxx xxxxxx xxxx xxxx xxx xxxxx
xxxxx xxxxxxxx xxx xxxxxxxx. Xxx xxxxxxx, xx xxx xxxxx xxxxx xxxxxx xxxx xxxx
xxxxxx, xx xxx xxxx xxxxxx xxxxxxxx xxx xxx xxxxx. Xxx xxxxxxx xxx xxxxx xx
Table~3.


\font\eri=cmti8

Xx xxxxxx xx xxxxxxxxxx xxxx xxxxx xxxxxx xxxx xxxxxxxx xxxx xxx xxx xxxxxxx
xxxxx xx xxx xxxx xxxxx xxx xx xxx xx xxx xxx xxxxxxxxx xx xxxxxxx xxxx xx xxx
xxx xxxxx xxxxx xxxxxx. Xxxxx~x xxxxxxx xxxx xxx xxxxx xxxxxxx, xx xx
xxxxxxxxx xxxxxxxxx xx xxxxxxxx xx xxxx xxxx xxxxxxx xxxx xxx xxxx xxx xxxx
xxxxxxxx xxxxxxxxxxx xxx xxx xxxxxx xx xx xxxxxxx. Xxxxxxxxx xxxxxxx xxxxx xxx
xx xxxxxxxx xxxxxx xxxx xxx xxxx xxxxxxx xx x xxxxx xxxxxxxxx, xxx xxx
xxxxxxxx xx xxx xxxxxxxxx? Xxx xxxx xxxxxxxx xxxxxxxxx xxxxxxxxx xxxxxxxxx xx
XX-xxx, xxx xxxxxxxxx xx xxx {xxx Xxxxxx xxxxxx Xxxx
xxx},\numberedfootnote{\erm Xxxx Xx xx xxx xxxx Xxxxxx Xxxxxxx xxxx xxxxx
xxxx.} ``X xxxxxxx xxxxx xxxxxxxxx xxx Xxxxxx xxx xxx xxxx'x xxxxx xx xxxxxxxx
xxxx xx xxxxxxxxx xxxxxx xxxxxx xxxx xxxxxx xxxx.'' Xxxx xxxxxx xxxxx xx x
xxxxxxxx xxxx xxxxxxxx xxxxx xxxxxx xxx xxxxxxxx xxx xxx xxxxxxx xxx xx xxxx
xxxxxxx xxx xxxxxx xxxxxx xxxxxxxx xxx xxxx xxxxxxxxxxx xxxx xxx xxx xxxxxxx
xxx xxxxxx xxxxxx xx Xxxxxx xxxxx xx xxxx, xxx xxxxx xxx xxx xxxxxxxx xx x
Xxxxxx xxxxxxx xxxx xxx xxxx? Xxx xxxxxx xxxxx xx xx xxxxxxxxx xxx xxxxxxx
xxxx xxx xxxx xxxx xx Xxxxx~x xxx xxxxxx xxxx xxx xxxxxxx xxxx xxxxx xx
xxxxxxx xx xxxxxx xxx xx xxxxx.

\singlecolumn

\medskip\bigbreak
\midinsert
\centerline{Table 3---Xxxxxxxxx Xxxx Xxxxxxxx xx Xxxxxxxx}
\smallbreak
\line \bgroup \hss
  \vbox to11\baselineskip{}
\hss \egroup
\endinsert

\bye
