\input eplain

This is some text\numberedfootnote{And this is the first footnote.} and
now some more text\numberedfootnote{The second.} and
finally\numberedfootnote{the third} some
more.\numberedfootnote{The fourth.} And again.

\footnotenumber = 0

\leftskip = 1in
Let's do another, this one should be numbered one
again\numberedfootnote{OK, here it is, numbered one}.  This paragraph
ends up being indented by one inch.  But this time the
footnote\numberedfootnote{This one should be like the others.} shouldn't be
indented.

\footnotemarkseparation = .16666em

\numberedlist
\li This is the first item in a list, and it has a
footnote\numberedfootnote{See? With a separation less than the others, too}.

This is the second paragraph of the first item, and it has
one\numberedfootnote{This is in the second paragraph.} too.
\endnumberedlist

\leftskip = 0pt
Now let's redefine the footnote command.
\let\footnote = \numberedfootnote

\everyfootnote = {\it\leftskip=1in}%

And do one\footnote{This is it, should be indented by one inch and in
italics.} footnote with that, which should also end up indented.

\vfill\eject

% Test the \footnoterule parameters.
% 
\footnoterulewidth = 6in
\footnoteruleheight = 4pt
\belowfootnoterulespace = 6pt

New page\footnote{This footnote should have a 4pt rule above it which 6 inches
wide and has 6 points of space below it.} to test the footnote rule
parameters.

\vfill\eject

\everyfootnote = {\bf}
\interfootnoteskip = 8pt
\footnoteruleheight = 0pt
\belowfootnoterulespace = 0pt
\skip\footins = 12pt

Some more\footnote{In bold, indented by the usual parindent, with no
extra space between it and the rule; in fact, no rule.} footnotes so we can
test\footnote{In bold, with 8pt extra space above it.} 
interfootnoteskip.

The big blank space following this will force a long footnote to be
broken across pages.

\vglue 41pc plus 1filll

This is the last line on the page\footnote{And this is the last
footnote; it should be like the last one. We are making this
footnote pretty long, so that it won't all fit on one page. I hope that
\TeX\ will break it across pages, so we see if the new footnote
formatting parameters still work in that case. Oh, boy, I can see we
should make this footnote really long; much longer than I can make up
garbage sentences like this for.  Well, maybe not.  I can be pretty
verbose when I have to be.  I suppose anyone can.  Did you notice I'm
inconsistent about leaving one space or two after sentence-ending
periods?  Punctuation, really.}, and it should have 12pt
(which is the footins skip) between it and the rule\footnote{I lied;
{\it this} is the last footnote}.

\end
