%From: "Michael Elphick" <M.J.Elphick@newcastle.ac.uk>
%Subject: Re: Eplain problems?
%To: karl@cs.umb.edu
%Date: Tue, 15 Jun 93 11:32:11 WET DST

% Here is the basic source for the (edited) version of the newsletter:
%--------------------------------------------------------------------------------

%  +-----------------------------------------------------+
%  |  The New Horse's Mouth:     Issue no 2 (May  1993)  |
%  +-----------------------------------------------------+

 \input nhm.style
 \presstrue
 \HORSE{2}

 \doublecolumns
 \EDITORIAL{An Editorial Apology}
   No doubt some of our readers have been wondering if the first
   number of the \NHM was just a flash in the pan!  We {\bf were}
   encouraged by the response to that first trial issue, and can
   only apologise for the long delay since then (for which the
   senior editor accepts responsibility).  Although we do say that
   this is an {\bf occasional} publication), we'll try and be a
   bit less so in future!
   In this issue, we start with an amusing account of a trip to
   Russia (arising unexpectedly from a student project), followed
   by some more local news (and our first cartoon), and finally
   reveal the answers to the Christmas Quiz!
 A line added in an attempt to reproduce the error noted by me;
 a line added in an attempt to reproduce the error noted by me;
 a line added in an attempt to reproduce the error noted by me;
 a line added in an attempt to reproduce the error noted by me;
 a line added in an attempt to reproduce the error HERE IT IS!

 \singlecolumn

 \BOXITEM{A Tale of Two CSSD Students in a Far-off Land}{
   Dick Snow has passed on to us the following account of the
   unexpected outcome of last year's CSSD Group Project.

    As some readers will know, the group project undertaken by the
 1991/92 class of M.Sc.~students in Computing Software and Systems
 Design (otherwise known as the CSSD students) was a study of the
 reliability and performance of international e-mail.  This project
 involved contact with a number of Computer Science departments and
 Computer Service departments in many -- often far-flung -- parts of
 the world.  In one case, so much interest was generated by their
 work, that two of them were invited to go to Russia to present a
 paper at a conference.  What follows can be loosely described as
 their "trip report".  It should perhaps be pointed out that this
 document did not constitute part of their submitted work in the
 assessment for their degree! }

 \doublecolumns

 ``Our story begins with a group of eight students on an advanced
 M.Sc degree course in Computing Software \& Systems Design.  This
 course is highly recommended for those with masochistic tendancies.
 It is only completed by those with an ability to work on a project
 for twenty-five hours out of every twenty-four whilst, at the same
 time, attending lots of very interesting(?) 9am meetings.  The
 purpose of these is to ensure that no student spend more than 3
 hours sleeping, and that everyone has far too much work to carry out
 in far too little time.

    Well, we managed to produce a project report!  It certainly will
 not go down in history as the greatest piece of English text ever to
 be written, but it was done whilst the authors were on sleep
 deprivation.  This is, of course, no excuse whatsoever as our
 project supervisors will doubtless verify.

    A group of Russian professors decided that this masterpiece was
 worthy of inclusion in a conference.  Incredible!  Well, that is
 wonderful, except for those who have to spend another 364 hours
 producing a paper!  This feat was carried out in record time and the
 result duly accepted by the organisers of CDS-92 (the first
 international conference that Russia has hosted, on Computer
 Distributed Systems).

    Next question - who's going?  There are several ways of answering
 this kind of question.  Unfortunately, the easiest solution
 terminates the lives of six of the group and is hence, not
 recommended.  Our solution was to bore six of the group into
 submission, leaving two of us to carry out the deed.

    Eventually we were set for Russia; cases packed, slides ready and
 a four-pack for the rail journey to London.  Everything perfect!
 Uh, not a chance!  Unfortunately, we did not account for the
 incredible standard of service offered by Russian airlines.  These
 converted world war two bombers are something else!  Those who have
 not had the pleasure of making a journey on one of these aeroplanes
 haven't lived.  The air hostesses wouldn't even help passengers
 remove their teeth from the seat in front.


    So here we are --- back in Newcastle.  Our trip was an amazing
 experience and we would like to thank all those concerned for making
 it possible; our project supervisors for their help and assistance,
 the department for our registration and accommodation fees, SERC for
 our travel expenses and last but not least, the remainder of the
 CSSD group.  They are to be thanked for being part of what has
 undoubtedly been the best and most successful CSSD group in the
 history of the course!''

 \rightline{{\bf Alex Jowett} and {\bf Simon Whittaker}}

    [{\bf N.B.} The serious side of the project is described in the
 CSSD students' interesting report entitled "An Analysis of
 International Electronic Mail Performance and Reliability" by
 E.~Diemoz et al., now available as Computing Science Technical
 Report No.~393]

 \BOXITEM{House For Sale}{

  Fed up with not being able to park?  Why not move house!

    3 Bedrooms, end terrace in quiet cul-de-sac.  Gardens front and
 rear.  Within walking distance of University, and on good bus
 routes in West Newcastle (near General Hospital and West Road shops).

 [For more details, contact Chris Ritson. (Ext 8175)]  }


 \ENDPAGE
 \bye

--------------------------------------------------------------------------------
