\vfil\eject
{\parindent=0pt\parskip=10pt\pretolerance=400
\frenchspacing
\centerline{\twelveb Bibliography}
\vskip 8pt

\def\Hpar{\par\hangindent .75in\hangafter=1}
\Hpar
{\csc American Mathematical Society} 
``Differences between \TeX82 and \TeX80.'' In {\sl \TeX\ and Metafont:
Errata and Changes}. Published for the \TeX\ Users Group by the American
Mathematical Society, Providence, RI. September 1983

\Hpar
{\csc Furuta, R., Scofield, J. {\rm and} Shaw, A.} ``Document Formatting
Systems: Survey, Concepts, and Issues.'' In {\sl Computing Surveys\/},
{\bf 14}, 3 (Sept.~1982), 417-472.

\Hpar
{\csc Kernighan, B.~W.~{\rm and} Cherry, L.~L.} ``Typesetting
Mathematics --- User's Guide (Second Edition).'' In {\sl Unix Programmer's
Manual\/}, Volume 2A.  Western Electric Co., Greensboro NC 1976

\Hpar
{\csc Knuth, D.~E.} ``\TeX, a system for technical text.'' In {\sl \TeX\ and
Metafont: New Directions in Typesetting\/}, part 2.  Digital Press
and the American Mathematical Society, Bedford, MA, and
Providence, RI.  1979

\Hpar
{\csc Knuth, D.~E.}
{\sl The \TeX book\/}.  Addison-Wesley Publishing and the
American Mathematical Society, Reading, MA, and Providence, RI.
1984

\Hpar
{\csc Lamport, L.}
{\sl The
L\kern-.2em\raise.3ex\hbox{\csc a}\kern-.09em\TeX\
Document Preparation System}.  Forthcoming in 1985.

\Hpar
{\csc Lesk, M.~E. {\rm and} Kernighan, B.~W.} ``Tbl --- A Program
to Format Tables.'' In {\sl UNIX Programmer's
Manual\/}, Volume 2A.  Western Electric Co., Greensboro NC 1976

\Hpar
{\csc Ossana, J.~F.} ``NROFF/TROFF User's Manual.'' In {\sl UNIX Programmer's
Manual\/}, Volume 2A.  Western Electric Co., Greensboro NC 1974

\Hpar
{\csc Reid, B.~K.} {\sl Scribe Introductory User's Manual\/}.  Unilogic
Inc., Pittsburgh PA, 1980

\Hpar
{\csc Urban, M.}{\sl An Introduction to
L\kern-.2em\raise.3ex\hbox{\csc a}\kern-.09em\TeX}.  TRW Inc., 1985

}
