%\vfil\eject
%\centerline{\twelveb Appendix B}
%\vskip 4pt
%\centerline{\twelveb Font Samples}
%\vskip 8pt
\Appendix{B}{Font Samples}
Plain \TeX\ predefines control sequences for the\ math\ italic, bold
and roman styles of the Computer Modern family in five-point,
seven-point and ten-point (the default).  The predefined fonts
are as follows:
$$\halign{#\quad&#\hfil\cr
|\tenrm|&Normal ten-point roman\cr
|\teni|&\teni ten-point\ math\ italic\cr
|\tenit|&\tenit ten-point text italic\cr
|\tenbf|&\tenbf ten-point bold extended\cr
|\tensl|&\tensl ten-point slanted roman\cr
|\tentt|&\tentt ten-point typewriter\cr
\noalign{\smallskip}
|\sevenrm|&\sevenrm seven-point roman\cr
|\seveni|&\seveni seven-point\ math\ italic\cr
|\sevenbf|&\sevenbf seven-point bold extended\cr
\noalign{\smallskip}
|\fiverm|&\fiverm five-point roman\cr
|\fivei|&\fivei five-point\ math\ italic\cr
|\fivebf|&\fivebf five-point bold extended\cr}$$

Note that the Math Italic fonts are intended to
be used in Mathematics Mode only, and hence
behave somewhat strangely if used for normal text

For fonts other than the predefined ones, the |\font| operation must be used
to associate a keyword with the font information file.
Here are some of the fonts distributed with the \Unix\
version of \TeX:
\vskip 10pt
\def\Fsample#1:#2.{\font\nonce=#1\line{\rlap{\tt#1}\hskip 1in\nonce#2\hfil}\vskip 3pt}
\Fsample amssq8:8-point sans-serif.
\Fsample amss10:10-point sans-serif.
\Fsample amssmc10:Emboldened 10-point sans-serif.
\Fsample amssmc40:40-point sans-serif.
\Fsample aminch:BIG.
\vskip 5pt
\Fsample amssqi8:8-point slanted sans-serif.
\Fsample amssi10:10-point slanted sans-serif.
\vskip 5pt
\Fsample amssbx10:10-point bold sans-serif.
\vskip 5pt
\Fsample amr6:6-point roman.
\Fsample amr8:8-point roman.
\Fsample amr9:9-point roman.
\Fsample amrc10:Narrower 10-point roman.
\vskip 5pt
\Fsample amsl8:8-point slanted.
\Fsample amsl9:8-point slanted.
\Fsample ambxsl10:10-point bold slanted.
\vskip 5pt
\Fsample ambx6:6-point bold.
\Fsample ambx8:8-point bold.
\Fsample ambx9:9-point bold.
\Fsample amb10:Narrower 10-point bold.
\Fsample ambc12:Narrower 12-point bold.
\vskip 5pt
\Fsample ammi6:6-point\ math\ italic.
\Fsample ammi8:8-point\ math\ italic.
\Fsample ammi9:9-point\ math\ italic.
\Fsample amti7:7-point text italic.
\Fsample amti8:8-point text italic.
\Fsample amti9:9-point text italic.
\Fsample ambi10:10-point\ bold\ math\ italic.
\vskip 5pt
\Fsample amtt8:8-point typewriter.
\Fsample amtt9:9-point typewriter.
\Fsample amsltt10:10-point slanted typewriter.
\Fsample amitt10:10-point italic typewriter.
\vskip 5pt
\Fsample amcsc10:10-point Small Capitals.
\vskip 5pt
\Fsample amu10:10-point unslanted italics.
\Fsample amdunh10:Tall thin letters for no known purpose.
\Fsample amnaro:Tall thin sans-serif letters.
