% -*-LaTeX-*-
% <BEEBE.TEX.DVI>IBMPC.LTX.1, 19-Aug-87 09:37:21, Edit by BEEBE

%------------------------------------------------------------------------
% EVERYTHING TO THE RIGHT OF A  %  IS A REMARK TO YOU AND IS IGNORED BY
% LaTeX.
%
% WARNING!  DO NOT TYPE ANY OF THE FOLLOWING 10 CHARACTERS AS NORMAL TEXT
% CHARACTERS:
%                &   $   #   %   _    {   }   ^   ~   \   
%
% The following seven are printed by typing a backslash in front of them:
%  \$  \&  \#  \%  \_  \{  and  \}.
%------------------------------------------------------------------------
\newcommand{\X}[1]{{#1}\index{{#1}}}
\documentstyle[11pt]{article}
\begin{document}
\pagestyle{empty}
\section*{\TeX{} DVI Driver Family---IBM PC Floppy Distribution}

With Release 2.08, the IBM PC floppy distribution overflowed the
6 floppy disks previously made with BACKUP and COPY, and it has
become necessary to reduce space by distributing compressed
source files in ARC format using the publicly-available ARC
utility.  A copy of ARC.EXE is included on the distribution.

The WC and LC subdirectories have been dropped from this
distribution, as they are no longer of interest.

The first three floppies are written with ARC and BACKUP.  To
restore them, do the following:

\begin{verbatim}
mkdir \DVI
mkdir \DVI\ARC
mkdir \DVI\DOC
:# The following will prompt for insertion of 
:# the 3 floppies in order
restore A: \DVI\ARC\*.*
:# restore the documentation directory
cd \DVI\DOC
\DVI\ARC\ARC x \DVI\ARC\DOC.ARC
:# restore the source directory
cd \DVI
\DVI\ARC\ARC x \DVI\ARC\DVI.ARC
DEL \DVI\ARC\*.ARC
\end{verbatim}

ARC has more capabilities than extraction; to see them, just run
the program with an empty command line.

Floppy \#4 contains DVIALW.EXE, DVIIMP.EXE, and DVIJEP.EXE, plus
XPORT.* for driving a serial printer with X-ON/X-OFF flow control
which MS-DOS does not support.

Floppy \#5 contains two versions of a Unix-like MAKE utility.
The top-level directory contains a version that works only on the
IBM PC; it is the one that I use to build the DVI family.  The
subdirectory PDMAKE contains a version that can be run on six
different operating systems; I use it on TOPS-20 and VAX VMS.

Use COPY to extract the files of interest from floppies \#4 and \#5.

\end{document}
