% -*-LaTeX-*-
% <BEEBE.TEX.DVI>IBMPC.LTX.1, 19-Aug-87 09:37:21, Edit by BEEBE

%------------------------------------------------------------------------
% EVERYTHING TO THE RIGHT OF A  %  IS A REMARK TO YOU AND IS IGNORED BY
% LaTeX.
%
% WARNING!  DO NOT TYPE ANY OF THE FOLLOWING 10 CHARACTERS AS NORMAL TEXT
% CHARACTERS:
%                &   $   #   %   _    {   }   ^   ~   \   
%
% The following seven are printed by typing a backslash in front of them:
%  \$  \&  \#  \%  \_  \{  and  \}.
%------------------------------------------------------------------------
\newcommand{\X}[1]{{#1}\index{{#1}}}
\documentstyle[11pt]{article}
\begin{document}
\pagestyle{empty}
\section*{\TeX{} DVI Driver Family---IBM PC Floppy Distribution [2.10]}

At Release 2.08, the IBM PC floppy distribution overflowed the
6 floppy disks previously made with BACKUP and COPY, and it has
become necessary to reduce space by distributing compressed
source files in ARC format using the publicly-available ARC
utility.  A copy of ARC.EXE is included on the distribution.

The floppies are written with BACKUP and contain mostly .ARC files.
To restore them, do the following:

\begin{verbatim}
:# Create directories if necessary 
:# (adapt names to your local conventions)
mkdir \dvi
mkdir \dvi\doc
mkdir \pcmake
mkdir \pdmake
:# The following will prompt for insertion of 
:# the floppies in order
restore A: *.*
:# The floppies are not needed for the following;

:# unpack the DVI documentation directory
cd \DVI\DOC
\ARC x \DOC.ARC
:# unpack the DVI source directory
cd \DVI
\ARC x \DVI.ARC
:# unpack PC make
cd \PCMAKE
\ARC x \PCMAKE.ARC
:# unpack .EXE files to your binary directory
cd \somewhere-in-PATH
\ARC x \DVIEXE.ARC
\end{verbatim}

ARC has more capabilities than extraction; to see them, just run
the program with an empty command line.

Executable versions of the DVI drivers are provided as DVIALW.EXE,
DVIIMP.EXE, and DVIJEP.EXE, plus XPORT.* for driving a serial printer
with X-ON/X-OFF flow control which MS-DOS does not support.

The distribution contains two versions of a Unix-like MAKE utility.
For the DVI development, I use the version in PCMAKE.  You should put
PCMAKE.EXE in your binary directory; rename it MAKE.EXE unless that
conflicts with a program you already have.  The version of Make in
PDMAKE.ARC supports more operating systems, and is used for the
TOPS-20 and VAX VMS DVI support; it can be unpacked in a similar
fashion.

DVI driver family version 2.10 has a number of new features, and is
now compiled with Microsoft C version 5.0.  See 00REVHST.TXT for
a change history, and 01README.TXT for details about use of the new
Microsoft C compiler.

\end{document}
